---
title: "A Preprocessing analysis of clinical data of TCGA-KIRP patients"
output: 
  html_document: 
    default
  github_document: 
    df_print: paged
    html_preview: FALSE
    keep_html: TRUE
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })    
---

This project contains a pipeline for analysis of The Cancer Genome Atlas Kidney - Renal Clear Cell Carcinoma (TCGA-KIRC) clinical data, from [Genomic Data Commons Data Portal](https://portal.gdc.cancer.gov/exploration?filters=%7B%22op%22%3A%22and%22%2C%22content%22%3A%5B%7B%22op%22%3A%22in%22%2C%22content%22%3A%7B%22field%22%3A%22cases.project.project_id%22%2C%22value%22%3A%5B%22TCGA-KIRC%22%5D%7D%7D%5D%7D) and [cBioPortal](https://www.cbioportal.org/study/summary?id=kirp_tcga).

In this section, the initial preprocessing is applied to clean the data and arrange following the Tidyverse philosophy. Exploratory Data Analysis summarizes their main characteristics. 


```{r, error=TRUE, message=FALSE, warning=FALSE, purl=FALSE, results='hide'}
## This chunk automatically generates a text .R version of this script when running within knitr.
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
try(knitr::purl(input,output,documentation=2,quiet=T), silent = TRUE)
# Avoid duplicate label error of knitr::purl
options(knitr.duplicate.label = 'allow')
# Code to browse the markdown file with renderized images.
knitr::opts_chunk$set(
  fig.path = "figs/render-"
)
```


```{r message=FALSE, warning=FALSE, echo = FALSE}
# Set the packages of interest
packages = c("tidyverse","skimr","finalfit")

# if a package is installed, it will be loaded
# otherwise, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})

suppressMessages(library("tidyverse"))
rm(packages)
setwd(".")
```


## 1. Data importing and visualizing

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
kirp_clin_raw <- read_delim("C:/Users/Ericd/Documents/KIRP Analisys/TCGA-CBIOPORTAL-DATA/kirp_tcga_pan_can_atlas_2018_clinical_data.tsv", "\t", 
                            escape_double = FALSE, 
                            trim_ws = TRUE)
```


```{r echo=FALSE, message=FALSE, results='hide', paged.print=TRUE}
class(kirp_clin_raw) 
dim(kirp_clin_raw) 
names(kirp_clin_raw) 
glimpse(kirp_clin_raw)
skim(kirp_clin_raw) 
#View(kirc_clin_raw)
```
```{r}
unique(kirp_clin_raw$`3q Status`)
```

## 2. Cleaning data

Select variables based on NA count (> 50% complete is a good choice!).
<!-- # TO DO @PATRICK: simplify code NA_sum? -->
<!-- # kirc_clean <- kirc_clin_raw %>% -->
<!-- #     summarise_all(~ sum(is.na(.)))  -->

```{r}
NA_fifty <- dim(kirp_clin_raw)[1]/2

NA_sum <- colSums(is.na(kirp_clin_raw))
NA_sum <- as.data.frame(NA_sum)
NA_sum <- tibble::rownames_to_column(NA_sum, "variables")
NA_sum <- NA_sum %>%
     filter(NA_sum < NA_fifty)

kirp_clean <- kirp_clin_raw %>%
     select(one_of(NA_sum$variables))
```

Remove duplicate observations:

```{r}
kirp_clean0 <- kirp_clean %>%
     distinct_at('Patient ID', .keep_all = TRUE)
```

Remove numeric variables with unique observations:  
<!-- # TO DO @PATRICK: function to select variables with unique observations? 
<!-- # kirc_cleanX <- kirc_clean1 %>% -->
<!-- #     summarise_if(is.numeric, ~ n=unique(.)) -->

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
kirp_clean0 %>%
     select_if(is.numeric) %>%
     skim()

kirp_clean1 <-  kirp_clean0  %>%
     select(!c('Number of Samples Per Patient', 
               'Last Alive Less Initial Pathologic Diagnosis Date Calculated Day Value'))
```

Remove character variables with unique observations:

```{r  message=FALSE, warning=FALSE, paged.print=TRUE}
kirp_clean1 %>%
  select_if(is.character) %>%
  skim()

kirp_clean2 <- kirp_clean1  %>%
     select(!c('Study ID',
               'Cancer Type',
               'Cancer Type Detailed',
               'TCGA PanCanAtlas Cancer Type Acronym',
               'ICD-10 Classification', 
               'International Classification of Diseases for Oncology, Third Edition ICD-O-3 Histology Code',
               'International Classification of Diseases for Oncology, Third Edition ICD-O-3 Site Code',
               'Informed consent verified',
               'Oncotree Code',
               'Sample Type',
               'Somatic Status',
               'Subtype',
               'Tumor Disease Anatomic Site',
               'Tumor Type',
               'Last Communication Contact from Initial Pathologic Diagnosis Date'))
```


Remove character variables with similar information - check each one!

```{r}
table(kirp_clean2$`Overall Survival Status`, exclude = NULL)
table(kirp_clean2$`Disease-specific Survival status`, exclude = NULL)

kirp_clean3 <- kirp_clean2  %>%
     select(!c('Sample ID',
               'Other Patient ID'))
```

Remove other variables not directly related to patient - check each one!

```{r}
kirp_clean4 <- kirp_clean3  %>%
     select(!c('American Joint Committee on Cancer Publication Version Type',
               'Center of sequencing',
               'Form completion date',
               '10p Status',
               '10q Status',
               '11p Status',
               '11q Status',
               '12p Status',
               '12q Status',
               '13 (13q) Status',
               '14 (14q) Status',
               '15 (15q) Status',
               '16p Status',
               '16q Status',
               '17p Status',
               '17q Status',
               '18p Status',
               '18q Status',
               '19p Status',
               '19q Status',
               '1p Status',
               '1q Status',
               '20p Status',
               '20q Status',
               '21 (21q) Status',
               '22 (22q) Status',
               '2p Status',
               '2q Status',
               '3p Status',
               '3q Status',
               '4p Status',
               '4q Status',
               '5p Status',
               '5q Status',
               '6p Status',
               '6q Status',
               '7p Status',
               '7q Status',
               '8p Status',
               '8q Status',
               '9p Status',
               '9q Status',
               'In PanCan Pathway Analysis',
               'Birth from Initial Pathologic Diagnosis Date',
               'Tissue Prospective Collection Indicator',
               'Tissue Retrospective Collection Indicator'))
```

## 3. Changing variables names

Using snake_style

NO NEED TO CHANGE VARIABLE NAMES
```{r}
kirp_clean4 <- kirp_clean4 %>%
     rename(patient_id = 'Patient ID',
            age = 'Diagnosis Age',
            aneuploidity_score = 'Aneuploidy Score',
            buffa_score = 'Buffa Hypoxia Score',
            metastasis_stg = 'American Joint Committee on Cancer Metastasis Stage Code',
            specific_survival_months = 'Months of disease-specific survival',
            specific_survival_status = 'Disease-specific Survival status',
            lymph_stg = 'Neoplasm Disease Lymph Node Stage American Joint Committee on Cancer Code',
            neoplasm_stg = 'Neoplasm Disease Stage American Joint Committee on Cancer Code',
            tumor_stg = 'American Joint Committee on Cancer Tumor Stage Code',
            disease_free_mth = 'Disease Free (Months)',
            disease_free_stt = 'Disease Free Status',
            survival_progress = 'Progress Free Survival (Months)',
            ethnicity = 'Ethnicity Category', 
            frac_genome_alter = 'Fraction Genome Altered',
            msi_score = 'MSI MANTIS Score',
            msi_sensor = 'MSIsensor Score',
            neoplasm_event = 'New Neoplasm Event Post Initial Therapy Indicator',
            neoadj_therapy = 'Neoadjuvant Therapy Type Administered Prior To Resection Text',
            prior_cancer = 'Prior Diagnosis',
            mutation_cnt = 'Mutation Count',
            over_surv_mth = 'Overall Survival (Months)',
            over_surv_stt = 'Overall Survival Status',
            race = 'Race Category',
            gender = 'Sex',
            tissue_site = 'Tissue Source Site',
            weight = 'Patient Weight',
            neoplasm_status = 'Person Neoplasm Cancer Status',
            progression_status = 'Progression Free Status',
            primary_lympn_node = 'Primary Lymph Node Presentation Assessment',
            radiation_therapy = 'Radiation Therapy',
            ragnum_score = 'Ragnum Hypoxia Score',
            winter_hypoxia = 'Winter Hypoxia Score'
            )
```

```{r}
summary(kirp_clean4)
```

## 4. Taming data

Use lubridate for dates

```{r}
kirp_clean4 <- kirp_clean4 %>%
     mutate_if(is.character, as.factor) %>%
     mutate(patient_id = as.character(patient_id),
            age = as.integer(age))
```

## 5. Checking NA patterns 

Check distincts types of NAs: MCAR, MAR, MNAR

```{r}
kirp_clean4  %>%
     missing_plot()

missing_glimpse(kirp_clean4)
```

## 6. Checking numeric variables

Check data distribution, plausible ranges, outliers;
Thinking about deleting outliers from dataset? Need to evaluate carefully each one!

```{r}
kirp_clean4 %>%
     select_if(is.numeric) %>%
     summary()
```

```{r}
ggplot(kirp_clean4, aes(age)) +
     geom_histogram(bins = 20, alpha = 0.8, color = "red")
```

YEAR_DIAGNOSE???
```{r}
ggplot(kirc_clean4, aes(year_diagnose)) +
     geom_histogram(bins = 20, alpha = 0.8, color = "red")
```

```{r}
ggplot(kirp_clean4, aes(x ='', y=survival_progress)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$survival_progress)
# error: disease_free_mth < 0
```

```{r}
ggplot(kirp_clean4, aes(x ='', y=frac_genome_alter)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$frac_genome_alter)
```

```{r}
ggplot(kirp_clean4, aes(x ='', y=aneuploidity_score)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$aneuploidity_score)
```

```{r}
ggplot(kirp_clean4, aes(x ='', y=mutation_cnt)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$mutation_cnt)
```

```{r}
ggplot(kirp_clean4, aes(x ='', y=over_surv_mth)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$over_surv_mth)
```


```{r}
ggplot(kirp_clean4, aes(x ='', y=buffa_score)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$buffa_score)
```


```{r}
ggplot(kirp_clean4, aes(x ='', y=winter_hypoxia)) +
     geom_boxplot(width = .5) +
     geom_jitter(width = 0.05, alpha = 0.2, color = "orange")
boxplot.stats(kirp_clean4$winter_hypoxia)
```
```{r}
summary('tumor_stg')
```


## 7. Checking categorical variables

Check frequency, lables and levels 

```{r}
kirp_clean4 %>%
     select_if(is.factor) %>%
     summary() 

# agregating levels
kirp_clin <- kirp_clean4 %>%
     mutate(tumor_stg = fct_collapse(tumor_stg,
                             T1 = c('T1', 'T1a', 'T1b'),
                             T2 = c('T2', 'T2a', 'T2b'),
                             T3 = c('T3', 'T3a', 'T3b', 'T3c')))

kirp_clin <- kirp_clin %>%
     mutate(prior_cancer = fct_collapse(prior_cancer, 
               Yes = c('Yes', 'Yes, History of Prior Malignancy', 'Yes, History of Synchronous/Bilateral Malignancy')))

kirp_clin <- kirp_clin %>%
     mutate(gender = fct_collapse(gender, Male = c('MALE', 'Male')))
                                        
kirp_clin <- kirp_clin %>%
     mutate(tissue_site = fct_collapse(tissue_site,
                         A = c('A3', 'AK', 'AS'),
                         B = c('B0', 'B2', 'B4', 'B8', 'BP'),
                         C = c('CJ', 'CW', 'CZ'),
                         OTHERS = c('G6', 'GK', 'MM', 'MW',
                                    '3Z', '6D', 'DV', 'EU', 'T7')))

# changing level names
kirp_clin <- kirp_clin %>%
     mutate(race = fct_recode(race, Asian='ASIAN', 'Black or African American'='BLACK OR AFRICAN AMERICAN', White='WHITE'))


kirp_clin %>%
     select_if(is.factor) %>%
     summary()
```

## 8. Correcting and checking again 

```{r}
# month values < 0
kirp_clin$disease_free_mth[kirp_clin$disease_free_mth == -11.79] <- 11.79
kirp_clin$disease_free_mth[kirp_clin$disease_free_mth == -0.62] <- 0.62

skim(kirp_clin)
```

## 9. Saving dataset

```{r}
write_csv(kirc_clin, path = "data/kirc_clin.csv")

rm(kirp_clean4, kirp_clean3, kirp_clean2, kirp_clean1, kirp_clean0, kirp_clean, NA_sum, NA_fifty)
```

## Further analysis

- [Comparison and Hyphotesis test](2.correlation.md) 
- [Logistic Regression Model](3.logistic_regression.md)

```{r}
sessionInfo()
```


